Ampersand uses Relation Algebra as a Programming Language

### 2018-01-27 2017 JLAMP/articleRamics.tex
Fixes for reviewer #1 completed

\title{Ampersand uses Relation Algebra as Programming Language}
\author[ou,ordina]{Stef Joosten\fnref{fn1}}
\ead{stef.joosten@ou.nl}
\address[ou]{Open Universiteit Nederland, Postbus 2960, 6401 DL Heerlen, the Netherlands}
\address[ordina]{Ordina NV,  Nieuwegein, the Netherlands}
\fntext[fn1]{ORCID 0000-0001-8308-0189}

\begin{abstract}
	Relation Algebra can be used as a programming language for building information systems.
	This paper demonstrates the principle by presenting a case study together with the theory behind programming in Relation Algebra.
	As a case study, we have developed a database application for legal reasoning.
	We discuss a small part of it to illustrate the mechanisms of programming in Relation Algebra.
	Beside being declarative, relation algebra comes with attractive promises for developing big software.
	The compiler that was used for this case study, Ampersand, is the result of an open source project.
	Ampersand has been tried and tested in practice and is available as free open source software\footnote{\tt https://github.com/AmpersandTarski/Ampersand}.
	This article is an extended version of a conference paper presented at RAMiCS-2017 \cite{JoostenRAMiCS2017}.
\end{abstract}

@@

	The author shares the conviction that a provably correct implementation of requirements is essential to the software development process~\cite{Boehm1981}.
	Even though formal methods are not widely used in computer science,
        Ampersand is built on formal foundations laid by specification methods such as Z~\cite{Z} and Alloy~\cite{Alloy2006}.
	In contrast with such methods,
	Ampersand is equipped with a software generator that generates information systems that consist of a database with user interfaces.
	Thus, specifying in Ampersand is developing software at the same time.

@@

	A rule is an equality between two terms.
	Terms are built from relations.
	Ampersand interprets every relation as a finite set of pairs, which are stored in the database.
	The phase in which Ampersand takes a script, and turns it into a database, is what we will refer to as \define{compile-time}.
	The phase in which a user interacts with the database, is what we will refer to as \define{run-time}.
	At run-time, Ampersand can decide which rules are satisfied by querying the database.
	The compiler generates all software needed to keep rules satisfied at run-time.
	If one of the rules is not satisfied as a result of data that has changed,
	that change is reverted (rolled back) to the original state.
	Either way, Ampersand ensures that all rules remain satisfied.
	Rules in Ampersand are kept satisfied rather than being executed directly.

@@

	Relations (from set $\rels$) are used in information systems to store data.
%	A \define{fact} is a statement that is true in a business context.
%	Facts are stored and kept as data in a computer.
	As data changes over time, so do the contents of these relations.
	In this paper relations are represented by variables $r$, $s$, and $d$.
	We represent the declaration of a relation $r$ by $\declare{nm}{A}{B}$,
	in which \id{nm} is a name and $A$ and $B$ are concepts.
	We call $A$ the source concept and $B$ the target concept of the relation.
	The pair $\pair{A}{B}$ is called the type of the relation.
	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concepts $A$ and $B$.

	The meaning of relations in Ampersand is defined by an interpretation function $\mathfrak{I}$.
	It maps each relation to a set of pairs.
	The type system of Ampersand guarantees that each pair satisfies:
\begin{equation}
	\pair{a}{b}\in\ti{\declare{nm}{A}{B}} \Rightarrow\ a \in A \wedge b \in B \label{typing of relations}
\end{equation}
	The type system is strong, which means that every atom has a type.
	The type system is static, which means that type checking is entirely done at compile-time.

@@

\section{Software for Legal Reasoning}
\label{sct:Conceptual analysis}
	As a case study, an argument assistance system, MirrorMe, was built.
	We have chosen to implement the ideas of Toulmin \cite{Toulmin1958},
	because his book ``The uses of Arguments'' is still one of the most influential works in the area of legal argument management%
\footnote{Thanks to Elleke Baecke for pointing us towards this source.}.
	Toulmin is regarded as the first scholar in modern history to come up with a usable theory of argumentation.
	Recent work typically draws on Toulmin, so his doctrine offers a good starting point.
	Toulmin's ideas have been implemented before, for example in a tool called ArguMed \cite{Verheij1999}.

	Software systems that support legal arguments have been around for many decades.
	Legal reasoning differs from logical reasoning because of the human judgments that are involved.
	Verheij~\cite{Verheij2003} distinguishes between argument assistance systems and automated reasoning systems.
	An automated reasoning system (ARS) mimics human reasoning in an attempt to automate a decision.
	An argument assisance system (AAS) helps its user to construct a decision.
	The literature on legal reasoning~\cite{Lind2007} makes it quite clear why mathematical logic alone does not suffice.
	Legal reasoning puts a human decision at the root of its notion of truth,
	whereas mathematical logic eradicates any human influence.
	Legal reasoning places human judgment at the core of the very notion of truth.
	For this reason, attempts to apply mathematical logic to legal judgments have had limited impact~\cite{Prakken2005}.
	Argument assistance systems~\cite{Verheij2005} have been more successful,
	because they respect the professional freedom of legal professionals to construct their own line of argumentation.
	They offer help in many different ways.
	For instance by looking up legal references, jurisdiction from the past, scholarly works etc.
	They can also help to construct and validate arguments by keeping arguments and evidence organized.
	They can store, disclose and share legal evidence.

	In software development, logic predominates.
	We have used it to build an AAS.
	The argumentation principles of Toulmin are \emph{implemented in} logic rather than \emph{replaced by} logic.
	The structure of MirrorMe was designed by conceptually analysing ideas of Toulmin,
	such as claim, warrant, argument, and rebuttal. They appear in MirrorMe as relations.
	The Ampersand compiler generates a conceptual data model
	to help the software developer to oversee all relations.
	Even though our case study yields a model that is a bit too large for this paper,
	we have printed a relevant part of it at the end of this paper in figure~\ref{fig:conceptual model}.
	This diagram gives an impression of what it looks like.
	It shows entities such as Statement, Argument, Binding, Template, Case, and Claim, together with their attributes.
	Attributes are the univalent and injective relations, which Ampersand derives from the script.
	Relations that are neither univalent nor injective are n-m relations between entities.
	In practice, the conceptual models have shown to be readable without much explanation for experienced database developers.

\section{Programming in Relation Algebra}
\label{sct:Programming in RA}
	This section illustrates the execution model of Ampersand by means of an example.
	A demonstration of this example was given at RAMiCS-2017.
	The example covers a tiny part of the whole MirrorMe program,
	which is enough to give a taste of software development in relation algebra.
	The use case for this example is the substitution of placeholders in text,
	for instance in replacing "\stmtText{The employee, [emp], is entitled to [increase]}" by
        "\stmtText{The employee, James, is entitled to 50 Euros.}".
	A user does such replacements when formulating arguments and reasoning with them in a legal case.
	In MirrorMe, a user can define a template with placeholders in square brackets.
	The strings \stmtText{[emp]} and \stmtText{[increase]} are called placeholders.
	The reason for using placeholders is that legal rules are stated in general terms, e.g.
	\stmtText{Every employee is entitled to an increase in salary}.
	The user of MirrorMe will pick a legal text (from a source he trusts) and replace parts of that text by placeholders.
	When the values of the placeholders are known, a statement can be made by substituting placeholders by actual phrases.
	When all substitutions are made, a statement remains.
	A \define{statement} is a phrase in natural language that can be either true or false.
	A \define{binding} assigns a piece of text to a placeholder.
	That text is the \define{phrase of} the binding and
	also the \define{value} of the placeholder.

@@

\begin{figure}[htb]
\begin{verbatim}
 ExecEngine run started
 ExecEngine satisfying rule 'signal phrase update'
 InsPair(resetS,Statement,Stat623,Statement,Stat623)
 ExecEngine satisfying rule 'flush substitutions'
 DelPair(substituted,Binding,Bind625,Statement,Stat623)
 ExecEngine satisfying rule 'reset statement text'
 InsPair(stmtShowText,Statement,Stat623,StmtText,
         The employee, [emp], is entitled to [increase].)
 ExecEngine satisfying rule 'done initializing'
 DelPair(resetS,Statement,Stat623,Statement,Stat623)
 ExecEngine satisfying rule 'substitute'
 InsPair(stmtShowText,Statement,Stat623,StmtText,
         The employee, James, is entitled to [increase].)
 InsPair(substituted,Binding,Bind625,Statement,Stat623)
 ExecEngine satisfying rule 'fill shownPhrase'
 InsPair(shownPhrase,Binding,Bind625,Phrase,James)
 ExecEngine run completed
\end{verbatim}
\caption{Log file of a substitution}
\label{fig:log file}
\end{figure}
	The log file of figure~\ref{fig:log file} has been taken from a computer that carries out the procedure to satisfy all rules.
	The machine will only act on rules that are violated.
	The first rule violated is ``signal phrase update'', in which the computer signals that something or someone has made a change
	in relation \id{phrase}.
	The procedure ends by doing the necessary substitutions,
	ensuring that all statements have the actual phrase of a placeholder in their text.

>>>>

\title{Ampersand uses Relation Algebra as a Programming Language}
\author[ou,ordina]{Stef Joosten\fnref{fn1}}
\ead{stef.joosten@ou.nl}
\address[ou]{Open Universiteit Nederland, Postbus 2960, 6401 DL Heerlen, the Netherlands}
\address[ordina]{Ordina NV,  Nieuwegein, the Netherlands}
\fntext[fn1]{ORCID 0000-0001-8308-0189}

\begin{abstract}
	Relation Algebra can be used as a programming language for building information systems.
	This paper demonstrates the principle by presenting a case study together with the theory behind programming in Relation Algebra.
	As a case study, we have developed a database application for legal reasoning.
	We discuss a small part of it to illustrate the mechanisms of programming in Relation Algebra.
	Beside being declarative, relation algebra comes with attractive prospects for developing big software.
	The compiler that was used for this case study, Ampersand, is the result of an open source project.
	Ampersand has been tried and tested in practice and is available as free open source software\footnote{\tt https://github.com/AmpersandTarski/Ampersand}.
	This article is an extended version of a conference paper presented at RAMiCS-2017 \cite{JoostenRAMiCS2017}.
\end{abstract}

@@

	The author shares the conviction that a provably correct implementation of requirements
	is essential to the software development process~\cite{Boehm1981}.
	Even though formal methods are not widely used in computer science,
        Ampersand is built on the foundations laid by formal specification methods such as Z~\cite{Z} and Alloy~\cite{Alloy2006}.
	In contrast with such methods,
	Ampersand is equipped with a software generator that generates information systems,
	each of which consists of a database with web-based user interfaces.
	Thus, specifying in Ampersand is developing software at the same time.

@@

	A rule is an equality between two terms.
	Terms are built from relations.
	Ampersand interprets every relation as a finite set of pairs, which are stored in the database.
	The phase in which Ampersand takes a script and turns it into a database, is what we will refer to as \define{compile-time}.
	The phase in which a user interacts with the database, is what we will refer to as \define{run-time}.
	At run-time, Ampersand can decide which rules are satisfied by querying the database.
	The compiler generates all software needed to keep rules satisfied at run-time.
	If one of the rules is not satisfied as a result of data that has changed,
	that change is reverted (rolled back) to the original state.
	Either way, Ampersand ensures that all rules remain satisfied.
	Rules in Ampersand are kept satisfied rather than being executed directly.

@@

	Relations (from set $\rels$) are used in information systems to store data.
%	A \define{fact} is a statement that is true in a business context.
%	Facts are stored and kept as data in a computer.
	As data changes over time, so do the contents of these relations.
	In this paper relations are represented by variables $r$, $s$, and $d$.
	We represent the declaration of a relation $r$ by $\declare{nm}{A}{B}$,
	in which \id{nm} is a name and $A$ and $B$ are concepts.
	We call $A$ the source concept and $B$ the target concept of the relation.
	The pair $\pair{A}{B}$ is called the \emph{type} of the relation.
	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concepts $A$ and $B$.

	The meaning of relations in Ampersand is defined by an interpretation function $\mathfrak{I}$.
	It maps each relation to a set of pairs.
	The type system of Ampersand guarantees that each pair in $r$ respects the type of $r$:
\begin{equation}
	\pair{a}{b}\in\ti{\declare{nm}{A}{B}} \Rightarrow\ a \in A \wedge b \in B \label{typing of relations}
\end{equation}
	The type system is strong, which means that every atom has a type.
	The type system is static, which means that type checking is entirely done at compile-time.

@@

\section{Software for Legal Reasoning}
\label{sct:Conceptual analysis}
	As a case study, an argument assistance system, MirrorMe, was built.
	We have chosen to implement the ideas of Toulmin \cite{Toulmin1958},
	because his book ``The uses of Arguments'' is still one of the most influential works in the area of legal argument management%
\footnote{Thanks to Elleke Baecke for pointing us towards this source.}.
	Recent work typically draws on Toulmin, so his doctrine offers a good starting point.
	Toulmin's ideas have been implemented before, for example in a tool called ArguMed \cite{Verheij1999}.

	Software systems that support legal arguments have been around for many decades.
	Legal reasoning differs from logical reasoning because of the human judgments that are involved.
	Verheij~\cite{Verheij2003} distinguishes between argument assistance systems and automated reasoning systems.
	An automated reasoning system (ARS) mimics human reasoning in an attempt to automate a decision.
	An argument assisance system (AAS) helps its user to construct a decision.
	The literature on legal reasoning~\cite{Lind2007} makes it quite clear why mathematical logic alone does not suffice.
	Legal reasoning puts a human decision at the root of its notion of truth,
	whereas mathematical logic eradicates any human influence.
	For this reason, attempts to apply mathematical logic to legal judgments have had limited impact~\cite{Prakken2005}.
	Argument assistance systems~\cite{Verheij2005} respect the professional freedom of each legal professional
	in constructing her own line of argumentation.
	AAS offer help in many different ways.
	For instance by looking up legal references, jurisdiction from the past, scholarly works etc.
	They can also help to construct and validate arguments by keeping arguments and evidence organized.
	They can store, disclose and share legal evidence.
	The MirrorMe system lets each user (whether a plaintiff, judge, or defendant) decide individually on the truth of every statement.
	Differences in truth assessments can then be made visible and manageable.

	Software development is founded on mathematical reasoning, whereas legal work is built on defeasible reasoning.
	So we have built an AAS in the mathematical formalism of relation algebra
	by \emph{implementing} Toulmin's principles%
\footnote{We deliberately did not want to squeeze legal reasoning into relation algebra, or any other objective formalism.}.
	The structure of MirrorMe was designed by conceptually analysing ideas of Toulmin,
	such as claim, warrant, argument, and rebuttal. They appear in MirrorMe as relations.
	The Ampersand compiler generates a conceptual data model
	to help the software developer to oversee all relations.
	Even though our case study yields a model that is a bit too large for this paper,
	we have printed a relevant part of it at the end of this paper in figure~\ref{fig:conceptual model}.
	This diagram gives an impression of what it looks like.
	It shows entities such as Statement, Argument, Binding, Template, Case, and Claim, together with their attributes.
	Attributes are the univalent and injective relations, which Ampersand derives from the script.
	Relations that are neither univalent nor injective are n-m relations between entities.
	In practice, the conceptual models have shown to be readable without much explanation for experienced database developers.

\section{Programming in Relation Algebra}
\label{sct:Programming in RA}
	This section illustrates the execution model of Ampersand by means of an example.
	A demonstration of this example was given at RAMiCS-2017.
	The example covers a tiny part of the whole MirrorMe program,
	which is enough to give a taste of software development in relation algebra.

	The use case for this example is the substitution of placeholders in text,
	for instance in replacing "\stmtText{The employee, [emp], is entitled to [increase]}" by
        "\stmtText{The employee, James, is entitled to 50 Euros.}".
	A user does such replacements when formulating arguments and reasoning with them in a legal case.
	In MirrorMe, a user can define templates.
	A \define{template} is a string in which substrings between square brackets serve as placeholders.
	In our example, the strings \stmtText{[emp]} and \stmtText{[increase]} are called placeholders.
	The reason for using placeholders is that legal rules are stated in general terms, e.g.
	\stmtText{Every employee is entitled to an increase in salary}.
	The user of MirrorMe will pick a legal text (from a source he trusts) and replace parts of that text by placeholders.
	When the values of the placeholders are known, a statement can be made by substituting placeholders by actual phrases.
	The substitution of a placeholder by a phrase is called a \define{binding}
	When all substitutions are made, a statement remains.
	A \define{statement} is a phrase in natural language that can be either true or false.
	In software, a statement is a template without placeholders.
	A binding assigns a piece of text to a placeholder.
	That text is the \define{phrase of} the binding and
	also the \define{value} of the placeholder.

@@

\begin{figure}[htb]
\begin{verbatim}
 ExecEngine run started
 ExecEngine satisfying rule 'signal phrase update'
 InsPair(resetS,Statement,Stat623,Statement,Stat623)
 ExecEngine satisfying rule 'flush substitutions'
 DelPair(substituted,Binding,Bind625,Statement,Stat623)
 ExecEngine satisfying rule 'reset statement text'
 InsPair(stmtShowText,Statement,Stat623,StmtText,
         The employee, [emp], is entitled to [increase].)
 ExecEngine satisfying rule 'done initializing'
 DelPair(resetS,Statement,Stat623,Statement,Stat623)
 ExecEngine satisfying rule 'substitute'
 InsPair(stmtShowText,Statement,Stat623,StmtText,
         The employee, James, is entitled to [increase].)
 InsPair(substituted,Binding,Bind625,Statement,Stat623)
 ExecEngine satisfying rule 'fill shownPhrase'
 InsPair(shownPhrase,Binding,Bind625,Phrase,James)
 ExecEngine run completed
\end{verbatim}
\caption{Log file of a substitution}
\label{fig:log file}
\end{figure}
	The log file of figure~\ref{fig:log file} has been taken from a computer that carries out the procedure to satisfy all rules.
	It illustrates how the machine is busy satisfying one rule after another by inserting or deleting elements in different relations.
	This example follows the event flow in a dependency graph, figure~\ref{fig:event flow}, which is discussed in section~\ref{sct:Programming}.
	The machine will only act on rules that are violated.
	The first rule violated is ``signal phrase update'', in which the computer signals that something or someone has made a change
	in relation \id{phrase}.
	The procedure ends by doing the necessary substitutions,
	ensuring that all statements have the actual phrase of a placeholder in their text.
